hormones

### 2020-01-17 analysis/07_hormones.Rmd
new hormone data

theme_B3 <- function () { 
  theme_classic(base_size = 11) +
    theme(
      panel.grid.major  = element_blank(),  # remove major gridlines
      panel.grid.minor  = element_blank(),  # remove minor gridlines
      plot.title = element_text(hjust = 0.5, face = "bold"), # center & bold 
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

@@

prolactin2 <- read_csv("../results/prolactin.csv") %>% 
          dplyr::rename("plasma_conc" = "prolactin_ng_mL",
                        "treatment" = "stage") %>% 
           dplyr::mutate(hormone = "prolactin") %>% 
           dplyr::mutate(treatment = fct_recode(treatment,
                            "m.n2" = "m.hatch",
                            "prolong" = "inc.prolong",
                            "extend" = "extend.hatch"),
                          study = fct_collapse(treatment,
                                 characterization = charlevels,
                                 manipulation = maniplevels1)) %>%

          dplyr::mutate(bird_id = gsub(".ATLAS", "" , bird_id)) %>% 
          dplyr::select(study, treatment, sex, bird_id, hormone, plasma_conc) 
head(prolactin2)

same <- inner_join(prolactin, prolactin2, by = "bird_id")
ggplot(same, aes(x = plasma_conc.x, y = plasma_conc.y, label = bird_id)) + geom_point() + geom_text()

all <- full_join(prolactin, prolactin2, by = "bird_id") 
onlyin1 <- anti_join(prolactin, prolactin2, by = "bird_id") 
onlyin2 <- anti_join(prolactin2, prolactin, by = "bird_id")

dim(prolactin)
dim(prolactin2)
dim(all)
dim(same)
dim(onlyin1)
dim(onlyin2)

print("only in old prolactin dataset") 
onlyin1$bird_id

print("only in new prolactin dataset") 
onlyin2$bird_id

prolctin1rnaseq <- full_join(prolactin, colData)
prolctin2rnaseq <- full_join(prolactin2, colData)

prolctin1rnaseq
prolctin2rnaseq
```

```{r wrangle-steroids}
PETC <- read_excel("../results/parental_care_hormone_RIA_data_master.xlsx", sheet = 2)  %>% 
                dplyr::mutate(treatment = fct_recode(stage,
                            "inc.d17" = "inc_d17",
                            "inc.d3" = "inc_d3",
                            "inc.d9" = "inc_d9",
                            "m.inc.d9" = "m_incd9",
                            "m.inc.d3" = "m_incd3",
                            "m.inc.d8" = "m_incd8",
                            "m.inc.d17" = "m_incd17",
                            "m.n2" = "m_hatch",
                            "control" = "stress_hpg")) %>%
                dplyr::filter(treatment %in% alllevels2) %>%   
                dplyr::mutate(sex = fct_recode(sex,
                            "female" = "f",
                            "male" = "m"),
                       study = fct_collapse(treatment,
                                characterization = charlevels,
                                manipulation = maniplevels1)) %>% 
              dplyr::mutate(bird_id = gsub("[[:punct:]]", "." , band_combo)) %>% 
              dplyr::select(study, treatment, sex, bird_id, hormone, plasma_conc) %>% 
             drop_na() %>%  droplevels()  
head(PETC)

@@

```{r correlations}
PETCP <- left_join(prolactin2, PETC, by = "bird_id") %>% drop_na()
PETCP$treatment.x <- factor(PETCP$treatment.x, levels = alllevels2)

plothormonecorrelations <- function(myhormone, myylab){
  PETCP %>% filter(hormone.y == myhormone) %>%
  ggplot(aes(x = plasma_conc.x, y = plasma_conc.y, color = treatment.x )) +
  geom_point() + 
  geom_smooth(method = "lm", se = F) +
  labs(x = "Prolactin (ng/mL)", y = myylab) +
  scale_color_manual(values = colorscharmaip) +
  facet_wrap(~treatment.x, nrow = 3) +
  theme(legend.position = "none")
}

plothormonecorrelations("cort", "Corticosterone (ng/mL")
plothormonecorrelations("estradiol", "Estradiol (ng/mL")
plothormonecorrelations("progesterone", "Progesterone (ng/mL")
plothormonecorrelations("testosterone", "Testosterone (ng/mL")
```

```{r combine-clean-split}
hormones <- rbind(prolactin, PETC)
hormones <- left_join(hormones, birds)

@@

hormones$okay <- ifelse(hormones$hormone == "cort" & hormones$plasma_conc > 30, "bad",
                    ifelse(hormones$hormone == "progesterone" & hormones$plasma_conc > 5, "bad", 
                           ifelse(hormones$hormone == "prolactin" & hormones$plasma_conc > 150, "bad", 
                        ifelse(hormones$hormone == "testosterone" & hormones$sex == "female", "bad",
                               ifelse(hormones$hormone == "estradiol" & hormones$sex == "male", "bad", "okay")))))
hormones <- hormones %>% filter(okay == "okay") %>% droplevels()
summary(hormones)
```

@@

```{r PRLonly}
hormonecharplot("prolactin", "PRL (ng/mL)")
```

@@

hormonecharplot("cort", "CORT (ng/mL)")
hormonecharplot("progesterone", "PROG (ng/mL)") 
d1 <- hormonecharplot("estradiol", "E (ng/mL)")
d2 <- hormonecharplot("testosterone", "T (ng/mL)")
plot_grid(d1,d2, nrow = 1)

@@

hormonemanipSteroids("cort", "CORT (ng/mL)", 10)
hormonemanipSteroids("progesterone", "PROG (ng/mL)",2.5)

@@

```{r aov-char}
prl.char <- hormones %>% filter(hormone == "prolactin", treatment %in% charlevels)   %>%  droplevels()
test.char <- hormones %>% filter(hormone == "testosterone", treatment %in% charlevels)   %>%  droplevels()
est.char <- hormones %>% filter(hormone == "estradiol", treatment %in% charlevels)   %>%  droplevels()
prog.char <- hormones %>% filter(hormone == "progesterone", treatment %in% charlevels)   %>%  droplevels()
cort.char <- hormones %>% filter(hormone == "cort", treatment %in% charlevels)   %>%  droplevels()

@@

aovSexTretment(prl.char, "PRL") # yes, sex difference (p = 0.00256), yes treatment effect (p < 2e-16), no interaction
aovSexTretment(cort.char, "CORT") # no sex difference (p = 0.779 ), small treatment effect (p = 0.0238),  no interaction
aovSexTretment(prog.char, "PROG") # no sex difference or treatment effect, signifiant interacion (p = 0.0107)

@@

```{r}
write.csv(hormones, "../results/hormones.csv", row.names = F)
write.csv(prolactin2, "../results/07_prolactin2.csv", row.names = F)

@@

rechelle <- hormones %>%
  dplyr::filter(hormone %in% c("prolactin", "cort"),
                treatment %in% c("lay", "inc.d9", "hatch", "n9")) %>%
  dplyr::select(-icons, -music, -iconpath, -okay)

rechelle$hormone <- factor(rechelle$hormone, levels = c("prolactin", "cort"))

prl.rechelle <- rechelle %>% filter(hormone == "prolactin")   %>%  droplevels()
cort.rechelle <- rechelle %>% filter(hormone == "cort")   %>%  droplevels()

@@

```{r sex}
hormones$hormone <- factor(hormones$hormone, 
                           levels = c("cort", "progesterone", "prolactin", 
                                      "estradiol", "testosterone"))

hormones %>%
  filter(study == "characterization") %>%
  ggplot(aes(x = sex, y = plasma_conc, fill = sex)) +
  geom_boxplot() +
  theme_B3() +
  facet_wrap(~hormone, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  labs(y = "concentration (ng/mL")

@@

printaovtablesex("cort", "CORT ~ sex")
printaovtablesex("progesterone", "Prog ~ sex")
printaovtablesex("prolactin", "PRL ~ sex")

TukeyHSD(aov(data = cort.char, plasma_conc ~ treatment), which = "treatment")

@@

hormones %>%
  filter(hormone == "cort") %>%
  group_by(sex) %>%
  summarise(mean = mean(plasma_conc),
            median = median(plasma_conc))

p1 <- hormones %>%
  filter(hormone == "cort",
         study == "characterization") %>%
  ggplot(aes(x = sex, y = plasma_conc, fill = sex)) +
  geom_boxplot() +
  theme_B3() +
  labs(y = "concentration (ng/mL)", subtitle = "Corticosterone" , x = NULL) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,8.5))
p1

cort <- hormones %>%
  filter(hormone == "cort",
         study == "characterization")

@@

p3 <- hormones %>%
  filter(hormone == "cort",
         treatment %in% c(controlsremoval, levelsremoval )) %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = treatment)) +
  geom_boxplot() +
  theme_B3() +
  scale_y_continuous(limits = c(0,8.5)) +
  scale_fill_manual(values = colorscharmaip) +
  theme(legend.position = "none") +
  labs( x = NULL, y = "concentration (ng/mL")
p3

p4 <- hormones %>%
  filter(hormone == "cort",
         treatment %in% c(controlstiming, levelstiming )) %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = treatment)) +
  geom_boxplot() +
  theme_B3() +
  scale_y_continuous(limits = c(0,8.5)) +
  scale_fill_manual(values = colorscharmaip) +
  theme(legend.position = "none") +
  labs(y = NULL, x = NULL)
p4

@@

```{r correlations2 }
hormoneswide <- hormones %>%
  select(study, treatment, sex, bird_id, hormone, plasma_conc) %>%
  pivot_wider( names_from = hormone, values_from  = plasma_conc, values_fn = list(plasma_conc = mean))

plotcorrelations <- function(i,j, myxlab, mylab, mysubtitle){
      p <- ggplot(hormoneswide, aes(x = i, y = j),na.rm = TRUE) +
       geom_point(aes(color = treatment), na.rm = TRUE) +
       geom_smooth( method = lm, color = "darkgrey", na.rm = TRUE ) +
       theme_B3() +
      #facet_wrap(~sex, scales = "free") +
        labs(x = myxlab, y = mylab, subtitle = mysubtitle) +
        theme(legend.position = "none") +
        scale_color_manual(values = colorscharmaip2)

@@

p1 <- plotcorrelations(hormoneswide$prolactin, hormoneswide$cort, "prolactin", "cort", "r = 0.221, p = 0.001")
p2 <- plotcorrelations(hormoneswide$prolactin, hormoneswide$progesterone, "prolactin", "progesterone", "r = -0.099, p = 0.2")
p3 <- plotcorrelations(hormoneswide$prolactin, hormoneswide$testosterone, "prolactin", "testosterone", "r = -0.258, p = 0.08")
p4 <- plotcorrelations(hormoneswide$prolactin, hormoneswide$estradiol, "prolactin", "estradiol", "r = 0.05, p = 0.6")

p5 <- plotcorrelations(hormoneswide$cort, hormoneswide$progesterone, "cort", "progesterone", "r = 0.104, p = 0.175")
p6 <- plotcorrelations(hormoneswide$cort, hormoneswide$estradiol, "cort", "estradiol", "r = -0.026, p = 0.8")
p7 <- plotcorrelations(hormoneswide$cort, hormoneswide$testosterone, "cort", "testosterone", "r = -0.108, p = 0.4")

p8 <- plotcorrelations(hormoneswide$progesterone, hormoneswide$estradiol, "progesterone", "estradiol", "r = 0.120, p = 0.3")
p9 <- plotcorrelations(hormoneswide$progesterone, hormoneswide$testosterone, "progesterone", "testosterone", "r = 0.133, p = 0.3")

allcorrs <- plot_grid(p1,p2,p3,p4,p5,p6,p7,p8,p9, nrow = 3, align = "hv", labels = "auto", label_size = 8)
allcorrs

plot_grid(p1 + theme(axis.text.x = element_text(angle = 0)),p3+ theme(axis.text.x = element_text(angle = 0))  + theme(legend.position = "right"), align = "hv", rel_widths = c(0.4,0.6))

cor.test(hormoneswide$prolactin, hormoneswide$cort)
cor.test(hormoneswide$prolactin, hormoneswide$progesterone)
cor.test(hormoneswide$prolactin, hormoneswide$estradiol)
cor.test(hormoneswide$prolactin, hormoneswide$testosterone)

cor.test(hormoneswide$cort, hormoneswide$estradiol)
cor.test(hormoneswide$cort, hormoneswide$testosterone)
cor.test(hormoneswide$cort, hormoneswide$progesterone)

cor.test(hormoneswide$progesterone, hormoneswide$estradiol)
cor.test(hormoneswide$progesterone, hormoneswide$testosterone)
```

>>>>

@@

@@

```{r wrangle-steroids}
PETC <- read_excel("../results/hormones.xlsx", sheet = 1)  %>% 
              dplyr::rename(corticosterone = cort, progesterone = p4,
                            estradiol = e2,  testosterone = t) %>% 
               dplyr::mutate(treatment = fct_recode(treatment...3,
                            "inc.d17" = "incd17",
                            "inc.d3" = "incd3",
                            "inc.d9" = "incd9",
                            "m.inc.d9" = "minc9",
                            "m.inc.d3" = "minc3",
                            "m.inc.d8" = "minc8",
                            "m.inc.d17" = "minc17",
                            "m.n2" = "m hatch",
                            "control" = "baseline")) %>%
                dplyr::filter(treatment %in% alllevels2) %>%   
                dplyr::mutate(sex = fct_recode(sex,
                            "female" = "f",
                            "male" = "m"),
                       study = fct_collapse(treatment,
                                characterization = charlevels,
                                manipulation = maniplevels1)) %>% 
              dplyr::mutate(bird_id = gsub("[[:punct:]]", "." , id)) %>% 
              dplyr::select(bird_id, treatment, sex, study, 
                            corticosterone, estradiol, testosterone, progesterone) %>% 
              pivot_longer(cols = corticosterone:progesterone,
                           names_to = "hormone", values_to = "plasma_conc",
                           values_drop_na = TRUE) %>% 
              drop_na() %>% droplevels()  
head(PETC)
```

```{r combine-clean-split}
hormones <- rbind(prolactin, PETC)

hormones$okay <- ifelse(hormones$hormone == "corticosterone" & hormones$plasma_conc > 30, "bad",
                    ifelse(hormones$hormone == "progesterone" & hormones$plasma_conc > 5, "bad", 
                           ifelse(hormones$hormone == "prolactin" & hormones$plasma_conc > 150, "bad", 
                        ifelse(hormones$hormone == "testosterone" & hormones$sex == "female", "bad",
                               ifelse(hormones$hormone == "estradiol" & hormones$sex == "male", "bad", "okay")))))
hormones <- hormones %>% filter(okay == "okay") %>% droplevels() %>% select(-okay)

write.csv(hormones, "../results/hormones.csv", row.names = F)

# make a winder one for correlations
hormoneswide <- hormones %>% pivot_wider(names_from = "hormone", values_from = "plasma_conc", 
                                         values_fn = list(plasma_conc = mean))
hormoneswide$treatment <- factor(hormoneswide$treatment, levels = alllevels)
head(hormoneswide)

# for facet wrapping
hormones <- left_join(hormones, birds)

@@

hormones$hormone <- factor(hormones$hormone, 
                           levels = c("prolactin", "corticosterone", "progesterone", 
                                      "estradiol", "testosterone"))
head(hormones)
```

@@

hormonecharplot("prolactin", "PRL (ng/mL)")
hormonecharplot("corticosterone", "CORT (ng/mL)")
hormonecharplot("estradiol", "e2 (ng/mL)")
hormonecharplot("testosterone", "t (ng/mL)")
hormonecharplot("progesterone", "p4 (ng/mL)")
```

@@

hormonecharplot("corticosterone", "CORT (ng/mL)")
hormonecharplot("progesterone", "PROG (ng/mL)") 
d1 <- hormonecharplot("estradiol", "E (ng/mL)")
d2 <- hormonecharplot("testosterone", "T (ng/mL)")
plot_grid(d1,d2, nrow = 1)

@@

hormonemanipSteroids("corticosterone", "CORT (ng/mL)", 10)
hormonemanipSteroids("progesterone", "PROG (ng/mL)",2.5)

@@

```{r aov-char}
prl.char <- hormones %>% filter(hormone == "prolactin", treatment %in% charlevels)   %>%  droplevels()
test.char <- hormones %>% filter(hormone == "testosterone", treatment %in% charlevels)   %>%  droplevels()
est.char <- hormones %>% filter(hormone == "estradiol", treatment %in% charlevels)   %>%  droplevels()
prog.char <- hormones %>% filter(hormone == "progesterone", treatment %in% charlevels)   %>%  droplevels()
cort.char <- hormones %>% filter(hormone == "corticosterone", treatment %in% charlevels)   %>%  droplevels()

@@

aovSexTretment(prl.char, "PRL") # yes, sex difference (p = 0.00256), yes treatment effect (p < 2e-16), no interaction
aovSexTretment(cort.char, "corticosterone") # no sex difference (p = 0.779 ), small treatment effect (p = 0.0238),  no interaction
aovSexTretment(prog.char, "PROG") # no sex difference or treatment effect, signifiant interacion (p = 0.0107)

@@

@@

rechelle <- hormones %>%
  dplyr::filter(hormone %in% c("prolactin", "corticosterone"),
                treatment %in% c("lay", "inc.d9", "hatch", "n9")) %>%
  dplyr::select(-icons, -music, -iconpath)

rechelle$hormone <- factor(rechelle$hormone, levels = c("prolactin", "corticosterone"))

prl.rechelle <- rechelle %>% filter(hormone == "prolactin")   %>%  droplevels()
cort.rechelle <- rechelle %>% filter(hormone == "corticosterone")   %>%  droplevels()

@@

```{r sex}
hormones %>%
  filter(study == "characterization") %>%
  ggplot(aes(x = sex, y = plasma_conc, fill = sex)) +
  geom_boxplot() +
  theme_B3() +
  facet_wrap(~hormone, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  labs(y = "concentration (ng/mL")

@@

printaovtablesex("prolactin", "PRL ~ sex")
```

@@

hormones %>%
  filter(hormone == "corticosterone") %>%
  group_by(sex) %>%
  summarise(mean = mean(plasma_conc),
            median = median(plasma_conc))

p1 <- hormones %>%
  filter(hormone == "corticosterone",
         study == "characterization") %>%
  ggplot(aes(x = sex, y = plasma_conc, fill = sex)) +
  geom_boxplot() +
  theme_B3() +
  labs(y = "concentration (ng/mL)", subtitle = "Corticosterone" , x = NULL) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0,8.5)) +
  scale_fill_manual(values = sexcolors)
p1

cort <- hormones %>%
  filter(hormone == "corticosterone",
         study == "characterization")

@@

p3 <- hormones %>%
  filter(hormone == "corticosterone",
         treatment %in% c(controlsremoval, levelsremoval )) %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = treatment)) +
  geom_boxplot() +
  theme_B3() +
  scale_y_continuous(limits = c(0,8.5)) +
  scale_fill_manual(values = colorscharmaip) +
  theme(legend.position = "none") +
  labs( x = NULL, y = "concentration (ng/mL")
p3

p4 <- hormones %>%
  filter(hormone == "corticosterone",
         treatment %in% c(controlstiming, levelstiming )) %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = treatment)) +
  geom_boxplot() +
  theme_B3() +
  scale_y_continuous(limits = c(0,8.5)) +
  scale_fill_manual(values = colorscharmaip) +
  theme(legend.position = "none") +
  labs(y = NULL, x = NULL)
p4

@@

```{r correlations2 }

plotcorrelations <- function(i,j, myxlab, mylab, mysubtitle){
      p <- ggplot(hormoneswide, aes(x = i, y = j), na.rm = TRUE) +
       geom_point(aes(color = treatment), na.rm = TRUE) +
       geom_smooth( method = lm, color = "darkgrey", na.rm = TRUE ) +
       theme_B3() +
      #facet_wrap(~sex, scales = "free") +
        labs(x = myxlab, y = mylab, subtitle = mysubtitle) +
        theme(legend.position = "none") +
        scale_color_manual(values = colorscharmaip2)

@@

hormoneswide

corrPRL <- hormoneswide %>%
  pivot_longer(cols = corticosterone:progesterone, names_to = "hormones", values_to = "conc", values_drop_na = TRUE) %>%
  filter(study == "characterization") %>%
  ggplot(aes(x = prolactin, y = conc)) +
  geom_point(aes(color = treatment)) +
  geom_smooth( color = "darkgrey", method = "lm") +
  facet_wrap(~hormones, nrow = 1, scales = "free_y")  +
  theme_B3() +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  scale_color_manual(values = colorscharmaip2) +
  guides(color = guide_legend(nrow = 1)) +
  labs(y = "conc. (ng/mL)", x = "prolactin (ng/mL)")
corrPRL

corrCORT <- hormoneswide %>%
  pivot_longer(cols = c(prolactin, estradiol, testosterone, progesterone), names_to = "hormones", values_to = "conc", values_drop_na = TRUE) %>%
  filter(study == "characterization") %>%
  ggplot(aes(x = corticosterone, y = conc)) +
  geom_point(aes(color = treatment)) +
  geom_smooth( color = "darkgrey", method = "lm") +
  facet_wrap(~hormones, nrow = 1, scales = "free_y")  +
  theme_B3() +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  scale_color_manual(values = colorscharmaip2) +
  guides(color = guide_legend(nrow = 1)) +
  labs(y = "conc. (ng/mL)", x = "corticosterone (ng/mL)")
corrCORT
```
